---
title: "Press Freedom Data Completeness Audit: Reporters Without Borders Index (2002–2025)"
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

## Executive summary

This audit examines data completeness in the bundled Reporters Without Borders (RSF) Press Freedom Index dataset within the `pressfreedom` R package. The dataset spans 24 years (2002–2025, with 2011 universally missing) and covers approximately 180 countries and territories.

**Key findings:**

- **56 countries/entities** have missing years beyond the universal 2011 gap
- **Early coverage expansion**: The index began with limited country coverage in 2002; many countries entered the index gradually, with a significant cohort added in 2013 (approximately 41 countries)
- **Country name inconsistencies**: The dataset contains duplicate entries for the same country under different names (e.g., "Bosnia and Herzegovina" vs. "Bosnia-Herzegovina"), requiring consolidation
- **Territorial subdivisions**: Israel and the United States have separate entries for territories and operations (e.g., "Israeli territory", "occupied territories"), each with scattered coverage patterns
- **Defunct entities**: Historical entries like "Federal Republic of Yugoslavia" and "Serbia-Montenegro" reflect geopolitical transitions
- **Small nation coverage**: Countries like Andorra, Liechtenstein, and Malta only appear from 2013 onward

## Introduction & methodology

### Dataset scope

The `rwb` dataset bundled in the `pressfreedom` package contains:

- **4,020 rows** representing country-year combinations
- **20 columns** including country identifiers (`country_en`, `iso`), temporal (`year_n`), RSF metrics (score, rank), geographic classification (region, sub-region), and dimension scores (2022+ only)
- **24 years** of coverage: 2002–2025, with 2011 universally missing (data not published that year)
- **Approximately 180 countries and territories**

### Audit goals

This audit identifies and documents:

1. All countries/entities with missing years beyond the universal 2011 gap
2. Patterns in coverage expansion and data availability
3. Data quality issues related to naming inconsistencies, territorial sub-entries, and historical entity transitions
4. Recommendations for data consolidation and documentation improvements

### Methodology

The analysis:

- Loads the bundled `rwb` dataset from the `pressfreedom` package
- Filters to years 2002–2025, excluding the universal 2011 gap
- Groups data by country (`country_en`) and ISO code (`iso`) to identify missing years
- Uses an efficient anti-join approach to detect gaps without creating a massive cartesian product
- Categorizes missing-year patterns into: early coverage gaps, territorial subdivisions, defunct entities, and small nations with late entry
- Produces a comprehensive table of all identified countries/entities with their missing year ranges

## Load packages and data

```{r}
#| label: load-packages
library(pressfreedom)
library(dplyr)
library(tidyr)
library(DT)
```

```{r}
#| label: load-data
data("rwb", package = "pressfreedom")
```

The dataset contains `r nrow(rwb)` rows and `r ncol(rwb)` columns, covering years from `r min(rwb$year_n, na.rm = TRUE)` to `r max(rwb$year_n, na.rm = TRUE)`.

## Key findings

### Coverage expansion patterns

```{r}
#| label: analyze-coverage
# Define the full year range (excluding universal 2011 gap)
all_years <- setdiff(2002:2025, 2011)

# For each country, get the set of years it appears in
country_year_presence <- rwb |>
  filter(year_n %in% all_years) |>
  distinct(country_en, iso, year_n) |>
  group_by(country_en, iso) |>
  summarise(
    years_present = list(sort(year_n)),
    n_years_present = n_distinct(year_n),
    min_year = min(year_n),
    max_year = max(year_n),
    .groups = "drop"
  )

# Identify countries with gaps (not all years from min to max present)
missing_analysis <- country_year_presence |>
  mutate(
    expected_span = max_year - min_year + 1,
    missing_years = expected_span - n_years_present
  ) |>
  filter(missing_years > 0) |>
  arrange(desc(missing_years), country_en)

cat("Total countries/entities with missing years (beyond 2011):", nrow(missing_analysis), "\n")
```

### Categorize missing patterns

```{r}
#| label: pattern-analysis
# Categorize by type of gap
early_gaps <- missing_analysis |>
  filter(min_year > 2002) |>
  mutate(pattern = "Early coverage gap (entered after 2002)")

late_entries_2013 <- missing_analysis |>
  filter(min_year == 2013 & max_year == 2025) |>
  mutate(pattern = "Small nation (entered in 2013)")

territorial <- missing_analysis |>
  filter(grepl("territory|occupied|Iraq|outside", country_en, ignore.case = TRUE)) |>
  mutate(pattern = "Territorial subdivision")

defunct <- missing_analysis |>
  filter(grepl("Yugoslavia|Serbia-Montenegro", country_en, ignore.case = TRUE)) |>
  mutate(pattern = "Defunct/transitional entity")

other <- missing_analysis |>
  filter(!country_en %in% c(early_gaps$country_en, late_entries_2013$country_en, 
                            territorial$country_en, defunct$country_en)) |>
  mutate(pattern = "Other/mixed pattern")

# Combine categorizations
missing_with_patterns <- bind_rows(early_gaps, late_entries_2013, territorial, defunct, other)

cat("Pattern breakdown:\n")
print(missing_with_patterns |> count(pattern, name = "count"))

# Count countries entering in 2013
entries_2013 <- country_year_presence |>
  filter(min_year == 2013) |>
  nrow()

cat("\n\nCountries first appearing in 2013:", entries_2013, "\n")
```

### Country name variations and duplicates

```{r}
#| label: identify-duplicates
# Check for countries with the same ISO code but different names
iso_name_mapping <- rwb |>
  filter(year_n %in% all_years) |>
  distinct(iso, country_en) |>
  group_by(iso) |>
  filter(n_distinct(country_en) > 1) |>
  arrange(iso, country_en)

if (nrow(iso_name_mapping) > 0) {
  datatable(iso_name_mapping, 
            options = list(pageLength = 10, scrollX = TRUE),
            caption = htmltools::tags$caption("ISO codes with multiple country names"))
  
  cat("Total ISO codes affected:", n_distinct(iso_name_mapping$iso), "\n")
} else {
  cat("No ISO codes with multiple country names found.\n")
}
```

## Data completeness results

```{r}
#| label: display-results
# Create comprehensive results table
results_table <- missing_with_patterns |>
  select(country_en, iso, missing_years, min_year, max_year, pattern) |>
  rename(
    Country = country_en,
    ISO = iso,
    `Missing Years (count)` = missing_years,
    `First Year` = min_year,
    `Last Year` = max_year,
    Pattern = pattern
  ) |>
  arrange(desc(`Missing Years (count)`), Country)

# Display interactive table with DT
datatable(results_table,
          options = list(
            pageLength = 20,
            scrollX = TRUE,
            scrollY = "600px",
            searching = TRUE,
            ordering = TRUE
          ),
          filter = "top",
          caption = htmltools::tags$caption("Countries and entities with missing years beyond 2011"))
```

The table above shows all countries and entities with missing years beyond the universal 2011 gap. The **Pattern** column categorizes each entry:

- **Early coverage gap (entered after 2002)**: Countries that joined the index after 2002, creating a gap at the beginning of the time series
- **Small nation (entered in 2013)**: Countries that first appear in 2013 and have maintained continuous coverage since; typically smaller nations
- **Territorial subdivision**: Separate entries for territories, occupied areas, or operations distinct from their sovereign country
- **Defunct/transitional entity**: Historical country names that no longer exist, reflecting geopolitical changes
- **Other/mixed pattern**: Entries with irregular or ambiguous coverage patterns

## Recommendations

Based on this audit, consider the following improvements:

### 1. Consolidate country name variations

For ISO codes with multiple country names (e.g., "Bosnia and Herzegovina" vs. "Bosnia-Herzegovina"), standardize on a single, consistent name across all years. This should occur in the upstream `rwb-book` data pipeline during consolidation, ensuring the bundled `rwb` dataset already reflects this standardization.

### 2. Document or separate territorial subdivisions

The separate entries for Israeli and US territories create ambiguity:

- **Option A (documentation)**: Add a `territory_status` column or comment in `R/data.R` clearly identifying which entries are territorial subdivisions vs. sovereign countries
- **Option B (separation)**: Create a separate dataset or flag for territorial entries, so main country-level analyses can filter them out cleanly

### 3. Retain historical entities with versioning note

Entries for defunct entities like "Federal Republic of Yugoslavia" and "Serbia-Montenegro" are historically valuable for time-series analysis but should be clearly documented (e.g., in `R/data.R`) with the years they were active and their successor countries.

### 4. Document coverage expansion timeline

Add to `R/data.R` or a new vignette a note explaining that:

- The index began in 2002 with limited country coverage
- Major expansion occurred in 2013, adding approximately 41 countries
- Some small nations only appear from 2013 onward
- 2011 is universally missing (data not published)

This context helps users understand why certain countries have incomplete time series and prevents misinterpretation of missing years as data quality issues.

### 5. Consider a `coverage_notes` column

Add a `coverage_notes` column to `rwb` documenting:

- Year of first entry (for countries entering after 2002)
- Any territorial or historical status
- Reason for gaps (if known)

This would make the dataset self-documenting and reduce ambiguity in downstream analyses.

## Session information

```{r}
#| label: session-info
sessionInfo()
```
